Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2024-01-04 analysis/paper/paper.tex
add raw data to package structure

%%% Version 3.4 Generated 2018/06/15 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%

@@

% Below is from frontiers
%
\bibliographystyle{frontiersinSCNS}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace}
%\doublespacing

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Dental calculus has proven to contain a wealth of
dietary information in the form of plant microfossils
(Henry and Piperno, 2008; Hardy et al., 2009),
proteins (Hendy et al., 2018; Warinner et al., 2014a),
and other organic residues (Buckley et al., 2014).
This dietary information can be preserved within the mineralised dental plaque
over many millennia, providing a unique window into the food-related behaviours of
past populations
(Henry and Piperno, 2008; Jovanović et al., 2021; Tao et al., 2020)
and extinct species (Henry et al., 2014; Hardy et al., 2012).

Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first
noted by Armitage (1975),
and later by Middleton and Rovner (1994),
and Fox and colleagues (1996). Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis (1997).\\
In more recent years, the study of dental calculus has increased exponentially,
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome
characterisation (including pathogens) through the analysis of DNA and proteins
(Adler et al., 2013; Warinner et al., 2014b),
microbotanical remains (Henry and Piperno, 2008; Hardy et al., 2009; Mickleburgh and Pagán-Jiménez, 2012),
other organic residues and proteins from dietary compounds
(Buckley et al., 2014; Hendy et al., 2018),
and nicotine use (Eerkens et al., 2018). Especially the extraction
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges
(Piperno and Dillehay, 2008; Henry et al., 2014; Tao et al., 2020; Jovanović et al., 2021).

Despite this, our knowledge of dental calculus and the incorporation pathways of
the various markers is limited (Radini et al., 2017), as is our
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested
by Tromp and colleagues (2017), specifically regarding
decalcification using HCl or EDTA.
The authors found significantly more starches with the EDTA extraction method
than the HCl extraction method; however, as noted by the authors, comparisons
involving archaeological calculus are problematic due to variability between and
within individuals.
Studies conducted on modern humans (Leonard et al., 2015)
and non-human primates (Power et al., 2015; Power et al., 2021)
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited,
despite meticulous documentation and observation, due to unknown variables and
uncertainty involved in this kind of \emph{in vivo} research. Dental calculus is a complex
oral biofilm with a multifactorial aetiology and variable formation rates both
within and between individuals (Jepsen et al., 2011; Haffajee et al., 2009),
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
within individuals (Froehlich et al., 1987; Nater et al., 2005),
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing (Hardy et al., 2018), as well as post-depositional
processes that are still being explored
(García-Granero, 2020; Mercader et al., 2018),
and it becomes clear that using
dental calculus to reconstruct diet is a highly unpredictable process.

@@

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

\hypertarget{biofilm-formation}{%
\subsection{Biofilm formation}\label{biofilm-formation}}

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues (1991) and
Shellis (1978). In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which
is autoclaved at 120°C, 1 bar overpressure, for 20 mins. The individual pegs
were the substrata on which the calculus grew. Using this system allowed for easy
transfer of the growing biofilm between saliva, feeding solutions,
and mineral solutions.

The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
described by Sissons and colleagues (1991).
It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778),
5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl\textsubscript{2},
5.2 mmol/l Na\textsubscript{2}HPO\textsubscript{4} (Sissons et al., 1991), 6.4 mmol/l NaHCO\textsubscript{3}
(Shellis, 1978), 2.5 mg/l haemin. This is subsequently
adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120°C,
1 bar overpressure), and supplemented with 5.8 \(\mu\)mol/l menadione, 5 mmol/l urea,
and 1 mmol/l arginine (Sissons et al., 1991).

@@

After 15 days, mineralisation was encouraged with a
calcium phosphate monofluorophosphate urea (CPMU) solution containing
20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l urea,
and (0.04 g/l MgCl)
(Pearce and Sissons, 1987; Sissons et al., 1991).
The substrata were submerged in 1 ml/well CPMU for six minutes, five times
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated
for 10 days until the end of the experiment on day 24
(see Figure \ref{fig:protocol-fig} for an overview of the protocol).

@@

\hypertarget{amylase-activity-detection}{%
\subsection{Amylase activity detection}\label{amylase-activity-detection}}

An \(\alpha\)-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14.
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4°C until the assay was conducted on day 18.
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing
H\textsubscript{2}O, while artificial saliva samples were compared against a standard curve
containing stock AS (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a
Multiskan FC Microplate Photometer (Thermo Scientific 51119000).
The protocol is a modified version of an Enzymatic Assay of \(\alpha\)-Amylase
(\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) (Bernfeld, 1955), which measures the amount of
maltose released from starch by \(\alpha\)-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 \(\mu\)mole of maltose in 6 minutes.
The detailed protocol can be found here: \url{https://www.protocols.io/view/amylase-activity-bw8jphun}.

\hypertarget{treatment-solutions}{%
\subsection{Treatment solutions}\label{treatment-solutions}}

@@

\hypertarget{extraction-method}{%
\subsection{Extraction method}\label{extraction-method}}

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 \(\tiny{M}\) ethylenediaminetetraacetic acid (EDTA)
(Tromp et al., 2017; Le Moyne and Crowther, 2021; Modi et al., 2020),
and vortexing for 3 days until the sample was completely dissolved.
Twenty \(\mu\)l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using
the pipette to ensure that the counted transects were representative of the
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material for more detail).

@@

\hypertarget{statistical-analysis}{%
\subsection{Statistical analysis}\label{statistical-analysis}}

Statistical analysis was conducted in R version 4.2.2 Patched (2022-11-10 r83330) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).

@@

Pearson's \emph{r} was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.
This was applied to total biofilm weight and starch count per mg
calculus (also z-score standardised) to account for differences in starch
concentration in the calculus (as per Wesolowski et al., 2010).

\hypertarget{results}{%
\section{Results}\label{results}}

@@

\hypertarget{no-amylase-activity-detected-in-the-model}{%
\subsection{No amylase activity detected in the model}\label{no-amylase-activity-detected-in-the-model}}

@@

\hypertarget{treatment-type-had-minimal-effect-on-biofilm-growth}{%
\subsection{Treatment type had minimal effect on biofilm growth}\label{treatment-type-had-minimal-effect-on-biofilm-growth}}

@@

\hypertarget{starch-counts}{%
\subsection{Starch counts}\label{starch-counts}}

@@

\hypertarget{proportion-of-available-starches-incorporated-in-samples}{%
\subsubsection{Proportion of available starches incorporated in samples}\label{proportion-of-available-starches-incorporated-in-samples}}

@@

\hypertarget{size-ratios-differ-between-solutions-and-samples}{%
\subsubsection{Size ratios differ between solutions and samples}\label{size-ratios-differ-between-solutions-and-samples}}

@@

\hypertarget{biofilm-weight-correlated-positively-with-extracted-starch-counts}{%
\subsubsection{Biofilm weight correlated positively with extracted starch counts}\label{biofilm-weight-correlated-positively-with-extracted-starch-counts}}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Here, we have provided a method for exploring the incorporation of dietary
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules
(of all sizes) present in the extracted samples were similar across all treatments
(0.06\% to 0.16\%),
despite large differences in absolute granule counts between wheat
(mean = 25,404,000)
and potato
(mean = 3,016,000)
solutions.\\
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat
and mixed solutions had the highest absolute mean count of starch granules, and
also had the highest absolute mean count of starch granules extracted from the
dental calculus
(Tables \ref{tab:solution-count-tbl} and \ref{tab:sample-count-tbl}).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and
degradation in the burial environment.
Despite the low proportion of granules recovered from the model calculus
(0.06\% to 0.16\%),
the absolute counts were still substantially greater than counts recovered from
archaeological remains
(Tromp and Dudgeon, 2015; Tromp et al., 2017; Wesolowski et al., 2010), which could in part be due to the
lack of oral amylase activity in our model.
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of
starch granules in dental calculus, and that starch granules extracted from dental
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues (2015) found individual
calculus samples to be a poor predictor of diet in a population, as many of the
consumed plants were missing from some individual samples, but were present in others.\\
Power and colleagues (2015)
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
(2015),
who found that plants producing starches 10--20 \(\mu\)m in
size were over-represented; however, the representation of granules larger than
20 \(\mu\)m in their study is unclear.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
maximum length were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, 174--176).
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(van de Velde et al., 2002; Jane et al., 1994; Reichert, 1913, 364--365; Seidemann, 1966, 174--176).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between size of biofilm deposit and
retained starch granules (Figure \ref{fig:cor-plot}A), meaning larger calculus
deposits contain a higher quantity of granules;
a result that contradicts findings from archaeological contexts
(Wesolowski et al., 2010; Dudgeon and Tromp, 2014).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot}B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = 5.53 mg) were on average larger than mixed treatment samples (mean = 4.28 mg) (Table \ref{tab:anova-tbl}); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

>>>>

%%% Version 3.4 Generated 2022/06/14 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%
%%%  NB logo1.jpg is required in the path in order to correctly compile front page header %%%

@@

% Below is from frontiers

@@

@@

\section{Introduction}\label{introduction}

Dental calculus has proven to contain a wealth of
dietary information in the form of plant microfossils
\citep{henryCalculusSyria2008, hardyStarchGranules2009},
proteins \citep{hendyProteomicCalculus2018, warinnerEvidenceMilk2014},
and other organic residues \citep{buckleyDentalCalculus2014}.
This dietary information can be preserved within the mineralised dental plaque
over many millennia, providing a unique window into the food-related behaviours of
past populations
\citep{henryCalculusSyria2008, jovanovicNeolithicCalculus2021, taoWheatCalculus2020}
and extinct species \citep{henryNeanderthalCalculus2014, hardyNeanderthalMedics2012}.

Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first
noted by Armitage \citeyearpar{armitageExtractionIdentification1975},
and later by Middleton and Rovner \citeyearpar{middletonOpalPhytoliths1994},
and Fox and colleagues \citeyearpar{foxPhytolithCalculus1996}. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis \citeyearpar{cummingsMayanCalculus1997}.\\
In more recent years, the study of dental calculus has increased exponentially,
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome
characterisation (including pathogens) through the analysis of DNA and proteins
\citep{adlerSequencingAncient2013, warinnerPathogensHost2014},
microbotanical remains \citep{henryCalculusSyria2008, hardyStarchGranules2009, mickleburghNewInsights2012},
other organic residues and proteins from dietary compounds
\citep{buckleyDentalCalculus2014, hendyProteomicCalculus2018},
and nicotine use \citep{eerkensDentalCalculus2018}. Especially the extraction
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges
\citep{pipernoStarchGrains2008, henryNeanderthalCalculus2014, taoWheatCalculus2020, jovanovicNeolithicCalculus2021}.

Despite this, our knowledge of dental calculus and the incorporation pathways of
the various markers is limited \citep{radiniFoodPathways2017}, as is our
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested
by Tromp and colleagues \citeyearpar{trompEDTACalculus2017}, specifically regarding
decalcification using HCl or EDTA.
The authors found significantly more starches with the EDTA extraction method
than the HCl extraction method; however, as noted by the authors, comparisons
involving archaeological calculus are problematic due to variability between and
within individuals.
Studies conducted on modern humans \citep{leonardPlantMicroremains2015}
and non-human primates \citep{powerChimpCalculus2015, powerRepresentativenessDental2021}
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited,
despite meticulous documentation and observation, due to unknown variables and
uncertainty involved in this kind of \emph{in vivo} research. Dental calculus is a complex
oral biofilm with a multifactorial aetiology and variable formation rates both
within and between individuals \citep{jepsenCalculusRemoval2011, haffajeeBiofilmPosition2009},
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
within individuals \citep{froehlichEffectOral1987, naterHumanAmylase2005},
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing \citep{hardyRecoveringInformation2018}, as well as post-depositional
processes that are still being explored
\citep{graneroStarchTaphonomy2020, mercaderExaggeratedExpectations2018},
and it becomes clear that using
dental calculus to reconstruct diet is a highly unpredictable process.

@@

\section{Materials and Methods}\label{materials-and-methods}

\subsubsection{Biofilm formation}\label{biofilm-formation}

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues \citeyearpar{sissonsMultistationPlaque1991} and
Shellis \citeyearpar{shellisSyntheticSaliva1978}. In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which
is autoclaved at 120°C, 1 bar overpressure, for 20 mins. The individual pegs
were the substrata on which the calculus grew. Using this system allowed for easy
transfer of the growing biofilm between saliva, feeding solutions,
and mineral solutions.

The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
described by Sissons and colleagues \citeyearpar{sissonsMultistationPlaque1991}.
It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778),
5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl\textsubscript{2},
5.2 mmol/l Na\textsubscript{2}HPO\textsubscript{4} \citep{sissonsMultistationPlaque1991}, 6.4 mmol/l NaHCO\textsubscript{3}
\citep{shellisSyntheticSaliva1978}, 2.5 mg/l haemin. This is subsequently
adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120°C,
1 bar overpressure), and supplemented with 5.8 \(\mu\)mol/l menadione, 5 mmol/l urea,
and 1 mmol/l arginine \citep{sissonsMultistationPlaque1991}.

@@

After 15 days, mineralisation was encouraged with a
calcium phosphate monofluorophosphate urea (CPMU) solution containing
20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l urea,
and (0.04 g/l MgCl)
\citep{pearceConcomitantDeposition1987, sissonsMultistationPlaque1991}.
The substrata were submerged in 1 ml/well CPMU for six minutes, five times
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated
for 10 days until the end of the experiment on day 24
(see Figure \ref{fig:protocol-fig} for an overview of the protocol).

@@

\subsubsection{Amylase activity detection}\label{amylase-activity-detection}

An \(\alpha\)-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14.
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4°C until the assay was conducted on day 18.
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing
H\textsubscript{2}O, while artificial saliva samples were compared against a standard curve
containing stock AS (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a
Multiskan FC Microplate Photometer (Thermo Scientific 51119000).
The protocol is a modified version of an Enzymatic Assay of \(\alpha\)-Amylase
(\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) \citep{bernfeldAmylase1955}, which measures the amount of
maltose released from starch by \(\alpha\)-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 \(\mu\)mole of maltose in 6 minutes.
The detailed protocol can be found here: \url{https://www.protocols.io/view/amylase-activity-bw8jphun}.

\subsubsection{Treatment solutions}\label{treatment-solutions}

@@

\subsubsection{Extraction method}\label{extraction-method}

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 \(\tiny{M}\) ethylenediaminetetraacetic acid (EDTA)
\citep{trompEDTACalculus2017, lemoyneCalculusPretreatments2021, modiCalculusMethodologies2020},
and vortexing for 3 days until the sample was completely dissolved.
Twenty \(\mu\)l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using
the pipette to ensure that the counted transects were representative of the
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material for more detail).

@@

\subsubsection{Statistical analysis}\label{statistical-analysis}

Statistical analysis was conducted in R version 4.3.2 (2023-10-31) \citep{Rbase} and
the following packages: tidyverse \citep{tidyverse2019}, broom \citep{Rbroom},
here \citep{Rhere}, and patchwork \citep{Rpatchwork}.

@@

Pearson's \emph{r} was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.
This was applied to total biofilm weight and starch count per mg
calculus (also z-score standardised) to account for differences in starch
concentration in the calculus \citep[as per][]{wesolowskiEvaluatingMicrofossil2010}.

\section{Results}\label{results}

@@

\subsubsection{No amylase activity detected in the model}\label{no-amylase-activity-detected-in-the-model}

@@

\subsubsection{Treatment type had minimal effect on biofilm growth}\label{treatment-type-had-minimal-effect-on-biofilm-growth}

@@

\subsubsection{Starch counts}\label{starch-counts}

@@

\paragraph{Proportion of available starches incorporated in samples}\label{proportion-of-available-starches-incorporated-in-samples}

@@

\paragraph{Size ratios differ between solutions and samples}\label{size-ratios-differ-between-solutions-and-samples}

@@

\paragraph{Biofilm weight correlated positively with extracted starch counts}\label{biofilm-weight-correlated-positively-with-extracted-starch-counts}

@@

\section{Discussion}\label{discussion}

Here, we have provided a method for exploring the incorporation of dietary
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules
(of all sizes) present in the extracted samples were similar across all treatments
(0.06\% to 0.16\%),
despite large differences in absolute granule counts between wheat
(mean = 25,404,000)
and potato
(mean = 3,016,000)
solutions.\\
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat
and mixed solutions had the highest absolute mean count of starch granules, and
also had the highest absolute mean count of starch granules extracted from the
dental calculus
(Tables \ref{tab:solution-count-tbl} and \ref{tab:sample-count-tbl}).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and
degradation in the burial environment.
Despite the low proportion of granules recovered from the model calculus
(0.06\% to 0.16\%),
the absolute counts were still substantially greater than counts recovered from
archaeological remains
\citep{trompDietaryNondietary2015, trompEDTACalculus2017, wesolowskiEvaluatingMicrofossil2010}, which could in part be due to the
lack of oral amylase activity in our model.
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of
starch granules in dental calculus, and that starch granules extracted from dental
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues \citeyearpar{leonardPlantMicroremains2015} found individual
calculus samples to be a poor predictor of diet in a population, as many of the
consumed plants were missing from some individual samples, but were present in others.\\
Power and colleagues \citeyearpar{powerChimpCalculus2015}
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
\citeyearpar{powerChimpCalculus2015},
who found that plants producing starches 10--20 \(\mu\)m in
size were over-represented; however, the representation of granules larger than
20 \(\mu\)m in their study is unclear.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
maximum length were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
\citep[174-176]{gismondiStarchGranules2019, haslamDecompositionStarch2004, seidemannStarchAtlas1966}.
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
\citetext{\citealp{vandeveldeStarchMorphology2002}; \citealp{janeAnthologyStarch1994}; \citealp[364-365]{reichertStarchBible1913b}; \citealp[174-176]{seidemannStarchAtlas1966}}.
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between size of biofilm deposit and
retained starch granules (Figure \ref{fig:cor-plot}A), meaning larger calculus
deposits contain a higher quantity of granules;
a result that contradicts findings from archaeological contexts
\citep{wesolowskiEvaluatingMicrofossil2010, dudgeonDietGeography2014}.
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot}B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = 5.53 mg) were on average larger than mixed treatment samples (mean = 4.28 mg) (Table \ref{tab:anova-tbl}); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

The mechanism by which starch granules are incorporated into plaque and calculus
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age
\citep{henryNeanderthalCalculus2014, buckleyDentalCalculus2014, wuDietEarliest2021}.
Studies have also shown that not all starch granules come from a dietary source.
Other pathways include cross-contamination from plant interactions in soil, such
as palm phytoliths adhering to the skin of sweet potatoes
\citep{trompDietaryNondietary2015},
or accidental ingestion not related to food consumption
\citep{radiniFoodPathways2017, radiniMedievalWomen2019}.\\
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to
tooth surfaces (including plaque and dental calculus) and mucous membranes
\citep{kashketFoodRetention1991, doddsCarbohydrateRetention1988}.
Bacteria also have the ability to adhere to starch granules
\citep{toppingResistantStarch2003},
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by \(\alpha\)-amylase \citep{kashketFoodParticles1996}.
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g., cooked) starch granules are more susceptible to enzymatic degradation,
while dehydrated starches will have a reduced susceptibility
\citep{lingstromStarchyFood1994, francoStarchDegradation1992, haslamDecompositionStarch2004, bjorckStarchProcessing1984, henryCookingStarch2009}.
Cracks on the surface of the dental calculus, as well as unmineralised islands
and channels may also be able to contain starch granules
\citep{powerSEMCalculus2014, tanCalculusUltrastructure2004, charlierSEMCalculus2010}.
Starch granules that are trapped in these pockets are (at least to some extent)
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus.
The size bias against large granules (\textgreater20 \(\mu\)m) from both wheat and potato
(Table \ref{tab:sample-prop-tbl}) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.
This was also suggested by Power and colleagues \citeyearpar{powerSEMCalculus2014}, who
observed clusters of starches within dental calculus, rather than an even
distribution across the surface of the dental calculus.
Granules trapped
in plaque/calculus may still be susceptible to hydrolysis, as \(\alpha\)-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
\citep{scannapiecoSalivaryAmylase1993, nikitkovaStarchBiofilms2013, tanBacterialViability2004, tanCalculusUltrastructure2004}.
After the death of an individual, starches within dental calculus are susceptible
to further degradation by post-depositional processes, depending on burial environment
(pH, temperature, moisture content, microorganisms)
\citep{henryCookingStarch2009, graneroStarchTaphonomy2020, haslamDecompositionStarch2004, francoStarchDegradation1992}.
Future study should explore how burial affects the recovery of starch from the
biofilm model.
